Political Polarization, Public Trust, and Healthcare Disparities: Unraveling COVID-19's Impact and Future Preparedness

### 2024-02-15 outputs/paper/paper.qmd
updated paper and llm

From all the data provided from @citearticle, @citedata1, and @citedata2, we mainly utilize  @data3, @data8, @data12, and @citedata2. The data from @citearticle is well-organized and require minimal cleaning for analysis.

The data collected from @data12 is at the county level as depicted on @tbl-sample. However, since our analysis focuses on people's political preferences at the state level, we create a new data frame called "vote_data_by_state" to aggregate voting data by state. We utilize the "group_by" function to generate new variables at the state level, calculating the total votes, total Republican party votes, total Democratic party votes, and the total vote difference between the two parties by county. Additionally, We also create variables for the average percentage of votes each party receives and the difference by each state.

@@

The data from @data2 lacks column names for each column. To retrieve data by column, we assign column names "State" and "Death_Rate" to the variables of interest. Subsequently, to find the top 10 states with the highest death rates, we rank the states based on death rate. Lastly, to create the final figure that incorporates the COVID death rate and people's political preferences by state, we merge the two datasets from @data12 and @citedata2 and proceed with our analysis.

@@

Incorporating this understanding into our analysis, our measurement for COVID death rates is drawn from a accumulated dataset last updated until March 2023. This temporal constraint means we would not be aware of the fluctuations of death rates over time, particularly during the peak of the pandemic in 2020. The data's primarily reliance on the reported statistics from hospitals, clinics, testing centres, and laboratories may introduces bias as excludes uncounted deaths from COVID. Especially during the critical period of health facilities capacity constraints,many individuals died outside of medical facilities without being documented. Therefore, our measurements for COVID death rates are only accounted for the documented death from COVID.

In addition,it's important to note that the death rate presented here reflects the entire population of a state. However, COVID-19 disproportionately affects older individuals compared to younger adults in the prime of their lives. As a result, this data does not capture the effect of COVID across different age groups. Instead, it only offers a general overview the state's population. In summary, we are unable to provide detailed insights into age-specific or temporal variations based on this dataset.

@@

Our measurement for each state's political preference is calculated from the 2020 election votes. There may be many scenarios that prevented eligible voters from voting. However, we are only able to assess the the state's political preference based on the people who voted, thereby potentially introducing selection bias. Additionally, issues such as ballot loss or voter disenfranchisement further compound the reliability of our measurements.

@@

Our results are summarized in the following figures. @fig-ELE illustrates the trend of life expectancy at birth across different racial groups over time. Unfortunately, due to the unavailability of data over time for Asian, American Indian, and Alaska Native communities, they have been excluded from the time series graph. @tbl-life-exp provides a more detailed breakdown of the life expectancy before and during COVID for Asian, American Indian, Alaska Native and other racial communities.

@@

Various studies have confirmed the existence of correlation between a state's political affiliation and its handling of COVID issues (@datz_2022). To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths that rank the top ten states with the highest death rates from COVID-19 per 100,000 people to examine the potential correlation between party preferences and COVID-related deaths. Notably, We found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates observed in Republican-leaning states may be attributed to their preference for less stringent measures compared to Democratic-leaning states, which tend to favor stricter measures (@vandusky-allen_shvetsova_2021).

@@

This begs the question as to why we are seeing these results. There isn't exactly a single answer to this question, however, we can certainly point out some considerable factors to this result. These factors can be broken down into the influence of political polarization, the impact of government transparency which ties into the effect it also had on public trust, and the role of social vulnerabilities and healthcare access disparities. We should also look at how to address these potential factors by implementing strategies in the future in regards to data collection and sharing. It is important to also note we were able to identify some weaknesses in our research that should be evaluated in the future.

@@

Investing in digital surveillance systems, which employ AI and machine learning to sift through diverse data sources for early outbreak detection, is essential for rapid response to health threats. An news article by `nature` [@citegoogle] talks about how Google has made an AI which claims to be "better bedside manner than human doctors — and makes better diagnoses". Furthermore, fostering public-private partnerships can harness the agility of the private sector and the public health expertise of governmental agencies to enhance data analytics capabilities.

@@

In our research, we identified several weaknesses in our data that is worth mentioning. The pandemic highlighted significant disparities in global surveillance, with countries possessing advanced healthcare systems paradoxically reporting higher COVID-19 mortality rates due to more accurate detection capabilities, underscoring the need for universal surveillance standards to accurately monitor diseases worldwide.

The United States, in particular, faced challenges with delayed data reporting, impeding swift public health responses and decision-making. The effects of this is self-evident as it makes us question if the data we are given is even correct. Furthermore, the often incomplete demographic data obscured the unequal effects of the pandemic on diverse communities, preventing targeted interventions for those most at risk, such as racial and ethnic minorities, thereby deepening existing health inequities.

@@

Enhancing data literacy among health professionals, policymakers, and the public will foster informed decision-making and trust in health interventions. Lastly, international collaboration in data sharing and analysis is crucial for the rapid dissemination of insights and best practices, bolstering global preparedness and response for future health crises.

>>>>

From all the data provided from @citearticle, @citedata1, and @citedata2, we mainly utilize  @data3, @data8, @data12, and @citedata2. The data from @citearticle is well-organized and requires minimal cleaning for analysis.

The data collected from @data12 is at the county level as depicted on @tbl-sample. However, since our analysis focuses on people's political preferences at the state level, we create a new data frame called "vote_data_by_state" to aggregate voting data by state. We utilize the "group_by" function to generate new variables at the state level, calculating the total votes, total Republican party votes, total Democratic party votes, and the total vote difference between the two parties by county. We also create variables for the average percentage of votes each party receives and the difference by each state.

@@

The data from @data2 lacks column names for each column. To retrieve data by column, we assign column names "State" and "Death_Rate" to the variables of interest. Subsequently, to find the top 10 states with the highest death rates, we rank the states based on death rate. Lastly, to create the final figure incorporating the COVID death rate and people's political preferences by state, we merge the two datasets from @data12 and @citedata2 and proceed with our analysis.

@@

Incorporating this understanding into our analysis, our measurement for COVID death rates is drawn from an accumulated dataset last updated until March 2023. This temporal constraint means we would not be aware of the fluctuations of death rates over time, particularly during the peak of the pandemic in 2020. The data's primarily reliance on the reported statistics from hospitals, clinics, testing centres, and laboratories may introduce bias as excludes uncounted deaths from COVID. Especially during the critical period of health facility capacity constraints, many individuals died outside of medical facilities without being documented. Therefore, our measurements for COVID death rates only accounted for the documented deaths from COVID.

In addition, it's important to note that the death rate presented here reflects the entire population of a state. However, COVID-19 disproportionately affects older individuals compared to younger adults in the prime of their lives. As a result, this data does not capture the effect of COVID across different age groups. Instead, it only offers a general overview of the state's population. In summary, we are unable to provide detailed insights into age-specific or temporal variations based on this dataset.

@@

Our measurement for each state's political preference is calculated from the 2020 election votes. There may be many scenarios that prevent eligible voters from voting. However, we can only assess the state's political preference based on the people who voted, thereby potentially introducing selection bias. Additionally, issues such as ballot loss or voter disenfranchisement further compound the reliability of our measurements.

@@

Our results are summarized in the following figures. @fig-ELE illustrates the trend of life expectancy at birth across different racial groups over time. Unfortunately, due to the unavailability of data over time for Asian, American Indian, and Alaska Native communities, they have been excluded from the time series graph. @tbl-life-exp provides a more detailed breakdown of the life expectancy before and during COVID-19 for Asian, American Indian, Alaska Native, and other racial communities.

@@

Various studies have confirmed the existence of a correlation between a state's political affiliation and its handling of COVID issues [@datz_2022]. To testify to the claim, we collected the voting data for each of the 50 states for the 2020 presidential election. We then selected the ten states with the highest proportions of Republican votes. These voting patterns are visualized in @fig-Vote, where red symbols represent Republican votes and blue symbols represent Democratic votes. Among the top ten states, Tennessee recorded the highest total number of votes, while Wyoming boasted the highest proportion of Republican votes.

Given the variation in population sizes among states, direct comparisons of COVID-19 case numbers are inherently flawed. As an alternative, we employed death rates per 100,000 people as a metric for evaluating each state's COVID preparedness and situation. We then produced @tbl-deaths that rank the top ten states with the highest death rates from COVID-19 per 100,000 people to examine the potential correlation between party preferences and COVID-related deaths. Notably, We found that six of the ten top Republican states made a reappearance in the top death rates table; these states are Oklahoma, West Virginia, Arkansas, Alabama, Tennessee, and Kentucky. The higher mortality rates observed in Republican-leaning states may be attributed to their preference for less stringent measures compared to Democratic-leaning states, which tend to favor stricter measures [@vandusky-allen_shvetsova_2021].

@@

This begs the question as to why we are seeing these results. There isn't exactly a single answer to this question, however, we can certainly point out some considerable factors to this result. These factors can be broken down into the influence of political polarization, the impact of government transparency which ties into the effect it also had on public trust, and the role of social vulnerabilities and healthcare access disparities. We should also look at how to address these potential factors by implementing strategies in the future regarding data collection and sharing. It is important to also note we were able to identify some weaknesses in our research that should be evaluated in the future.

@@

Investing in digital surveillance systems, which employ AI and machine learning to sift through diverse data sources for early outbreak detection, is essential for rapid response to health threats. A news article by `nature` [@citegoogle] talks about how Google has made an AI that claims to be "better bedside manner than human doctors — and makes better diagnoses". Furthermore, fostering public-private partnerships can harness the agility of the private sector and the public health expertise of governmental agencies to enhance data analytics capabilities.

@@

In our research, we identified several weaknesses in our data that are worth mentioning. The pandemic highlighted significant disparities in global surveillance, with countries possessing advanced healthcare systems paradoxically reporting higher COVID-19 mortality rates due to more accurate detection capabilities, underscoring the need for universal surveillance standards to accurately monitor diseases worldwide.

The United States, in particular, faced challenges with delayed data reporting, impeding swift public health responses and decision-making. The effects of this are self-evident as it makes us question if the data we are given is even correct. Furthermore, the often incomplete demographic data obscured the unequal effects of the pandemic on diverse communities, preventing targeted interventions for those most at risk, such as racial and ethnic minorities, thereby deepening existing health inequities.

@@

Enhancing data literacy among health professionals, policymakers, and the public will foster informed decision-making and trust in health interventions. Lastly, international collaboration in data sharing and analysis is crucial for the rapid dissemination of insights and best practices, bolstering global preparedness and response to future health crises.
